exp1_lik_by_trial

### 2020-08-13 Bayesian_models_exp1.Rmd
added plots lik by trial

library(dplyr)

@@

# dat <- subset(dat,human_id == "QSuzBXpbyRc370HsAACW")

@@

# if human generalizes over games, we can run update the prior generally
# if human does not generalize, we need to reset the prior at the start of each ?stage?
exp1_Bayes_model_LL <- function(par,data, generalize = c("game","no")) {
  # data is a subset for a single participant
  generalize <- match.arg(generalize)
  alpha <- par[1] # probability that ai opponent plays according to strategy
  if(length(par) > 1) {
    beta <- par[2] # probability that human plays according to best response 
  } else{
    beta <- 1 # probability that human plays according to best response 
  }

@@

  ret <- -2*sum(dat$loglik)
  if(is.infinite(ret) || is.nan(ret)) {
    return(1e+300)
  } else {
    return(ret)
  }
  # return(-2*sum(dat$loglik))
}

@@

```{r}
# exp1_Bayes_model_LL(0.4,data = dat, generalize = "game")
```

>>>>

@@

dat <- subset(dat,human_id == "QSuzBXpbyRc370HsAACW")

@@

# if human generalizes over games, we can run update the prior generally
# if human does not generalize, we need to reset the prior at the start of each ?stage?
exp1_Bayes_model_LL <- function(par,data, generalize = c("game","no"), return_value = "-2loglik") {
   # return_value=c("-2loglik","likelihood_by_trial")
  # data is a subset for a single participant
  generalize <- match.arg(generalize)
  alpha <- par[1] # probability that ai opponent plays according to strategy
  if(length(par) > 1) {
    beta <- par[2] # probability that human plays according to best response 
  } else{
    beta <- 1 # probability that human plays according to best response 
  }

@@

  ret <- -2*sum(dat$loglik)
  if(return_value == "-2loglik") {
    if(is.infinite(ret) || is.nan(ret)) {
    return(1e+300)
    } else {
      return(ret)
    } 
  } else if ((return_value == "likelihood_by_trial")) {
    return(exp(dat$loglik))
  }

}

@@

```{r}
exp1_Bayes_model_LL(0.4,data = dat, generalize = "no", "likelihood_by_trial")
```

### 2020-08-13 Report/draft_report_v1.Rmd
added plots lik by trial

Next we compared the performance of players whose actions are consistent with each of our hypothesized models. Figure\ \@ref(fig:exp1-cumScores) shows the average cumulative performance of players across games, for participants grouped by which model best fits their behavior in experiment 1. We can see that participants whose actions are most consistent with learning a ToM opponent model in a Bayesian way had the best overall performance (both with and witout transfer), followed by Q-learning conditional on last round play. EWA, QL and random players had, understandably the lowest performance.

>>>>

Next we compared the performance of players whose actions are best fit by each of our hypothesized models. Figure\ \@ref(fig:exp1-cumScores) shows the average cumulative performance of players across games, for participants grouped by which model best fits their behavior in experiment 1. We can see that participants whose actions are most consistent with learning a ToM opponent model in a Bayesian way had the best overall performance (witout transfer), followed by Q-learning conditional on last round play. EWA, QL and random players had, understandably the lowest performance.

### 2020-08-13 comp_model_1019.Rmd
added plots lik by trial

BIC_weights <- comp_BICs["ID"]
BIC_weights[,2:ncol(comp_BICs)] <- t(apply(comp_BICs[,2:ncol(comp_BICs)], 1, function(i) exp(-0.5*(i-min(i)) )))
colnames(BIC_weights) <-colnames(comp_BICs)

BIC_weights[,2:ncol(BIC_weights)] <- t(apply(BIC_weights[,-1], 1, function(i) round(i/sum(i),2)))

>>>>

exp1_BIC_weights <- comp_BICs["ID"]
exp1_BIC_weights[,2:ncol(comp_BICs)] <- t(apply(comp_BICs[,2:ncol(comp_BICs)], 1, function(i) exp(-0.5*(i-min(i)) )))
colnames(exp1_BIC_weights) <-colnames(comp_BICs)

exp1_BIC_weights[,2:ncol(exp1_BIC_weights)] <- t(apply(exp1_BIC_weights[,-1], 1, function(i) round(i/sum(i),2)))

### 2020-08-13 exp1_lik_by_trial.Rmd
added plots lik by trial

---
title: "Bayesian_models_exp2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

@@

dat <- read.csv("data20180719.csv")

####### test data

dat <- subset(dat,human_id == "QSuzBXpbyRc370HsAACW")

#######

dat <- as_tibble(dat) %>% group_by(human_id,game)
dat <- dat %>%
  mutate(ai_action_prev = lag(ai_action,1), human_action_prev = lag(human_action,1))

```{r}
## the following computes probabilities of ai actions assuming
## strategy is always followed. (level refers to ai level)

@@

```{r}
# group data by subjective component id

# if human generalizes over games, we can run update the prior generally
# if human does not generalize, we need to reset the prior at the start of each ?stage?
exp1_Bayes_model_LL <- function(par,data, generalize = c("game","no"), return_value = "-2loglik") {
   # return_value=c("-2loglik","likelihood_by_trial")
  # data is a subset for a single participant
  generalize <- match.arg(generalize)
  alpha <- par[1] # probability that ai opponent plays according to strategy
  if(length(par) > 1) {
    beta <- par[2] # probability that human plays according to best response 
  } else{
    beta <- 1 # probability that human plays according to best response 
  }

  prior <- c(1,1,1) # prior alpha for dirichlet on p(level)
  prior <- prior/sum(prior)

  # use alpha to change the "deterministic" prediction

  dat <- ungroup(data) %>%
     mutate_at(.vars = vars(c(starts_with("pred_a1"),starts_with("pred_a2"),starts_with("pred_a3"))), 
                            funs (case_when(
      game == "numbers" ~  alpha*. + (1-alpha)/5,
                  TRUE ~ alpha*. + (1-alpha)/3))) %>%
     mutate_at(.vars = vars(c(starts_with("pred_a4"),starts_with("pred_a5"))), 
                            funs (case_when(
      game == "numbers" ~  alpha*. + (1-alpha)/5,
                  TRUE ~ 0)))

  if(generalize == "no") {
    # group by game and stage
    dat <- group_by(dat,game,add=TRUE)
  }

  # compute likelihood of ai action
  dat <- dat %>%
    mutate(
      lik_level0 = case_when(
        ai_action_num == 1 ~ pred_a1_level0,
        ai_action_num == 2 ~ pred_a2_level0,
        ai_action_num == 3 ~ pred_a3_level0,
        ai_action_num == 4 ~ pred_a4_level0,
        ai_action_num == 5 ~ pred_a5_level0
      ),
      lik_level1 = case_when(
        ai_action_num == 1 ~ pred_a1_level1,
        ai_action_num == 2 ~ pred_a2_level1,
        ai_action_num == 3 ~ pred_a3_level1,
        ai_action_num == 4 ~ pred_a4_level1,
        ai_action_num == 5 ~ pred_a5_level1
      ),
      lik_level2 = case_when(
        ai_action_num == 1 ~ pred_a1_level2,
        ai_action_num == 2 ~ pred_a2_level2,
        ai_action_num == 3 ~ pred_a3_level2,
        ai_action_num == 4 ~ pred_a4_level2,
        ai_action_num == 5 ~ pred_a5_level2
      )
    )

  # use likelihood to compute the posterior predictive probability of each level
  dat <- dat %>%
    mutate(logpost_level0 = lag(log(prior[1]) + cumsum(log(lik_level0)),default=log(prior[1])),
           logpost_level1 = lag(log(prior[2]) + cumsum(log(lik_level1)),default=log(prior[2])),
           logpost_level2 = lag(log(prior[3]) + cumsum(log(lik_level2)),default=log(prior[3]))) %>%
    # you can add or subtract any constant from the log-likelihoods; this can aid in precision
    mutate(min = pmin(logpost_level0,logpost_level1, logpost_level2)) %>%
      mutate(normalize = exp(logpost_level0 - min) + exp(logpost_level1 - min) + exp(logpost_level2 - min)) %>%
        mutate(post_level0 = exp(logpost_level0 - min)/normalize,
               post_level1 = exp(logpost_level1 - min)/normalize,
               post_level2 = exp(logpost_level2 - min)/normalize)

   #cat(as.character(c(round(dat$post_level0,2), round(dat$post_level1,2), round(dat$post_level2,2)), "\n", "\n"))

  # use posterior predictive probability to predict probability of each ai and then human action
  # opp is ai and self is human player here
  dat <- dat %>%
    mutate(p_opp_a1 = post_level0*pred_a1_level0 + post_level1*pred_a1_level1 + post_level2*pred_a1_level2,
           p_opp_a2 = post_level0*pred_a2_level0 + post_level1*pred_a2_level1 + post_level2*pred_a2_level2,
           p_opp_a3 = post_level0*pred_a3_level0 + post_level1*pred_a3_level1 + post_level2*pred_a3_level2,
           p_opp_a4 = post_level0*pred_a4_level0 + post_level1*pred_a4_level1 + post_level2*pred_a4_level2,
           p_opp_a5 = post_level0*pred_a5_level0 + post_level1*pred_a5_level1 + post_level2*pred_a5_level2
    ) %>%
      mutate(p_self_a1 = case_when(
              game == "numbers" ~ (1-beta)*(1/3) + beta*p_opp_a5,
              TRUE ~ (1-beta)*(1/3) + beta*(p_opp_a3)
                ),
             p_self_a2 = case_when(
               game == "numbers" ~ (1-beta)*(1/3) + beta*p_opp_a1,
               TRUE ~ (1-beta)*(1/3) + beta*(p_opp_a1)
             ),
             p_self_a3 = case_when(
               game == "numbers" ~ (1-beta)*(1/3) + beta*p_opp_a2,
               TRUE ~ (1-beta)*(1/3) + beta*(p_opp_a2)
             ),
             p_self_a4 = case_when(
               game == "numbers" ~ (1-beta)*(1/3) + beta*p_opp_a3,
               TRUE ~ 0
             ),
             p_self_a5 = case_when(
               game == "numbers" ~ (1-beta)*(1/3) + beta*p_opp_a4,
               TRUE ~ 0
             )
        )

  # now finally compute the likelihood of human actions 
  dat <- dat %>% 
    mutate(
      loglik = case_when(
        human_action_num == 1 ~ log(p_self_a1),
        human_action_num == 2 ~ log(p_self_a2),
        human_action_num == 3 ~ log(p_self_a3),
        human_action_num == 4 ~ log(p_self_a4),
        human_action_num == 5 ~ log(p_self_a5))
    )

  #cat(as.character(exp(dat$loglik)), "\n", "\n")

  ret <- -2*sum(dat$loglik)
  if(return_value == "-2loglik") {
    if(is.infinite(ret) || is.nan(ret)) {
    return(1e+300)
    } else {
      return(ret)
    } 
  } else if ((return_value == "likelihood_by_trial")) {
    return(exp(dat$loglik))
  }

}

@@

```{r}
exp1_Bayes_model_LL(0.4,data = dat, generalize = "no", "likelihood_by_trial")
```

```{r}
#PLot likelihood for first participant as function of parameter

dat <- subset(dat, human_id == "QSuzBXpbyRc370HsAACW")
likelihoods <- list()
for (t in 1:95) {
  likelihoods[t] <- exp1_Bayes_model_LL(t/100,data = dat, generalize = "game")
}

jpeg(file="QSuzBXpbyRc370HsAACW.jpeg")
plot(unlist(likelihoods))
dev.off()

```{r, ECHO = FALSE}

library(DEoptim)

exp1_Bayes_game_Tr <- exp1_Bayes_no_Tr <- list()

# # USing DEOptim 
for(id in levels(dat$human_id)) {
  tdat <- subset(dat,human_id == id)
  ctrl <- DEoptim.control(NP = 20, itermax=50,parallelType = 1, packages = c("dplyr"))
  exp1_Bayes_game_Tr[[id]] <- DEoptim(exp1_Bayes_model_LL, lower=c(0), upper = c(1), data = tdat,
                                  generalize = "game",control=ctrl)
  exp1_Bayes_no_Tr[[id]] <- DEoptim(exp1_Bayes_model_LL, lower=c(0), upper = c(1), data = tdat,
                                  generalize = "no",control=ctrl)
}

# using optim

# for(id in levels(dat$human_id)) {
#   tdat <- subset(dat,human_id == id)
#   # ctrl <- DEoptim.control(NP = 20, itermax=50)
#   
#   Bayes_same_game[[id]] <- optim(c(0.1),fn=Bayes_model_LL,data=tdat, distinct_opponent = FALSE, 
#                                  generalize = "game", lower = c(0), upper = c(1),method="L-BFGS-B")
#   
#   
#   Bayes_same_stage[[id]] <- optim(c(0.1),fn=Bayes_model_LL,data=tdat, distinct_opponent = FALSE, 
#                                  generalize = "stage", lower = c(0), upper = c(1),method="L-BFGS-B")
# 
#   Bayes_same_no[[id]] <- optim(c(0.1),fn=Bayes_model_LL,data=tdat, distinct_opponent = FALSE, 
#                                  generalize = "no", lower = c(0), upper = c(1),method="L-BFGS-B")
# 
#   Bayes_distinct_game[[id]] <- optim(c(0.1),fn=Bayes_model_LL,data=tdat, distinct_opponent = TRUE, 
#                                  generalize = "game", lower = c(0), upper = c(1),method="L-BFGS-B")
# 
#   Bayes_distinct_stage[[id]] <- optim(c(0.1),fn=Bayes_model_LL,data=tdat, distinct_opponent = TRUE, 
#                                  generalize = "stage", lower = c(0), upper = c(1),method="L-BFGS-B")
# 
#   Bayes_distinct_no[[id]] <- optim(c(0.1),fn=Bayes_model_LL,data=tdat, distinct_opponent = TRUE, 
#                                  generalize = "no", lower = c(0), upper = c(1),method="L-BFGS-B")
# }
```
```{r}

exp1_Bayes_game_Tr[["QSuzBXpbyRc370HsAACW"]]$optim$bestval
exp1_Bayes_game_Tr[["QSuzBXpbyRc370HsAACW"]]$optim$bestmem[[1]]
```

>>>>

---
title: "exp1_lik_by_trial"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
```

```{r}
# Get Data for experiment 1, remove unecessary columns

dat <- read.csv("exp1_data.csv")
#####
# dat <- subset(dat, human_id == "38VxtUSv_h6RR5-tAAA2")
####

optim_results <- read.csv("All_results.csv")

@@

#preparing data for functions to claculate likelihood

# dat <- as_tibble(dat) %>% group_by(human_id,game)
dat <- dat %>%
  mutate(ai_action_prev = lag(ai_action,1), human_action_prev = lag(human_action,1))

@@

```{r}
# All_results <- data.frame()
# for(id in unique(dat$human_id)) {
#   All_results <- rbind(All_results,
#                        data.frame(
#                          "ID" = id,
#                          "condition" = dat[dat$human_id==id,"condition"][1],
#                          "Random_BIC" = -2*(100*log(1/3) + 50*log(1/5)),

lik_by_trial <- data.frame()

for(id in unique(dat$human_id)) {

  tdat <- dat %>% subset(human_id == id)

  # QLS likelihoods 
  pars_QLS <- unlist(filter(optim_results, ID == id)%>%select(QL_states_alpha,QL_states_Beta))
  tdat$QLS_lik <- Q_learn_states(pars_QLS,tdat,"likelihood_by_trial",gamma = 0)

  # Bayes model no transfer likelihoods
  theta_NT <- unlist(filter(optim_results, ID == id)%>%select(theta_no_transfer))
  tdat$Bayes_NT_lik <- exp1_Bayes_model_LL(theta_NT ,data = tdat, generalize = "no", "likelihood_by_trial")

  # Bayes model with Transfer likelihood
  theta_Tr <- unlist(filter(optim_results, ID == id)%>%select(theta_transfer))
  tdat$Bayes_Tr_lik <- exp1_Bayes_model_LL(theta_Tr ,data = tdat, generalize = "game", "likelihood_by_trial")

  lik_by_trial <- rbind(lik_by_trial,tdat)
}
```

```{r}
library(ggplot2)

mean_lik <- lik_by_trial %>%
  select(human_id,condition,game, round,QLS_lik,Bayes_NT_lik,Bayes_Tr_lik) %>%
  group_by(condition,game,round) %>%
  summarise(mean_QLS_lik = mean(QLS_lik, na.rm = TRUE),
            mean_NT_lik = mean(Bayes_NT_lik, na.rm = TRUE),
            mean_Tr_lik = mean(Bayes_Tr_lik, na.rm = TRUE))

mean_lik$game <- factor(mean_lik$game,levels=c("rps","fwg","numbers"))

data_long <- gather(mean_lik, strategy, probability, mean_QLS_lik:mean_Tr_lik, factor_key=TRUE)

# make sure the different games are ordered in the way they were played
data_long$game <- factor(data_long$game,levels=c("rps","fwg","numbers"))
ggplot(data_long,aes(x=round,y=probability,colour=strategy)) + geom_line() + facet_grid(game~condition)

@@
